Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2023-12-14 main.tex
:pencil: Add RCG method

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
% \newacronym{af}{AF}{Amplify-and-Forward}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}

@@

\begin{section}{Point-to-Point MIMO}
	\begin{subsection}{Channel Power Maximization}
		Consider a \gls{bd} \gls{ris} with $N_\mathrm{S}$ elements, which is divided into $G$ groups of equal $L$ elements.
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\left\lVert \mathbf{H}^\mathrm{D} + \sum_g\nolimits \mathbf{H}_g^\mathrm{B} \mathbf{\Theta}_g \mathbf{H}_g^\mathrm{F} \right\rVert _\mathrm{F}^2}{\label{op:channel_power}}{}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}_L, \quad \forall g \in \mathcal{G} \triangleq \{1,\ldots,G\}}{}{}
		\end{maxi!}
		For \emph{symmetric} BD-RIS, the problem has been solved in
		\begin{itemize}
			\item Matteo's paper \cite{Nerini2023}: SISO and equivalent\footnote{Single-stream MIMO with given precoder and combiner.};
			\item Ignacio's paper \cite{Santamaria2023}: SISO and directless MISO/SIMO.
		\end{itemize}

		\begin{remark}
			% The difficulty of \eqref{op:channel_power} is that the \gls{ris} needs to balance the (additive) direct-indirect eigenspace alignment and the (multiplicative) forward-backward eigenspace alignment.
			The difficulty of \eqref{op:channel_power} is that the \gls{ris} needs to balance the additive (direct-indirect) and multiplicative (forward-backward) eigenspace alignment.
			% align the direct and indirect eigenspaces while preserving the
			Interestingly, it has the same form as the \emph{weighted orthogonal Procrustes problem} \cite{Gower2004}:
			\begin{mini!}
				{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{C} - \mathbf{A \Theta B} \rVert _\text{F}^2}{\label{op:weighted_orthogonal_procrustes}}{}
				\addConstraint{\mathbf{\Theta}^\mathsf{H} \mathbf{\Theta}=\mathbf{I}}{}{}
			\end{mini!}
			There exists no trivial solution to \eqref{op:weighted_orthogonal_procrustes}.
			One lossy transformation, by moving $\mathbf{\Theta}$ to one side \cite{Bell2003}, formulates a standard orthogonal Procrustes problem:
			\begin{mini!}
				{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{A}^\dagger \mathbf{C} - \mathbf{\Theta B} \rVert _\text{F}^2}{\label{op:standard_orthogonal_procrustes}}{}
				\addConstraint{\mathbf{\Theta}^\mathsf{H} \mathbf{\Theta}=\mathbf{I}}{}{}
			\end{mini!}
			\eqref{op:standard_orthogonal_procrustes} has a global optimal solution $\mathbf{\Theta}^\star = \mathbf{U} \mathbf{V}^\mathsf{H}$, where $\mathbf{U}$ and $\mathbf{V}$ are left and right singular matrix of $\mathbf{\mathbf{A}^\dagger \mathbf{C} \mathbf{B}^\mathsf{H}}$ \cite{Golub2013}.
			This low-complexity solution will be compared with the one proposed later.
		\end{remark}

		Inspired by \cite{Nie2017}, we propose an iterative algorithm to solve \eqref{op:channel_power}.
		The idea is to successively approximate the quadratic objective with a sequence of affine functions and solve the resulting subproblems in closed form.

		\begin{proposition}
			Start from any $\mathbf{\Theta}^{(0)}$, the sequence
			\begin{equation}
				\mathbf{\Theta}_g^{(r+1)} = \mathbf{U}_g^{(r)} \mathbf{V}_g^{(r)}, \quad \forall g \in \mathcal{G}
			\end{equation}
			converges to a stationary point of \eqref{op:channel_power}, where $\mathbf{U}_g^{(r)}$ and $\mathbf{V}_g^{(r)}$ are left and right singular matrix of
			\begin{equation}
				\begin{split}
					\mathbf{M}_g^{(r)}
					& = {\mathbf{H}_g^\mathrm{B}}^\mathsf{H} \mathbf{H}^\mathrm{D} {\mathbf{H}_g^\mathrm{F}}^\mathsf{H} + \sum_{g' < g} {\mathbf{H}_{g'}^\mathrm{B}}^\mathsf{H} \mathbf{H}_{g'}^\mathrm{B} \mathbf{\Theta}_{g'}^{(r+1)} \mathbf{H}_{g'}^\mathrm{F} {\mathbf{H}_{g'}^\mathrm{F}}^\mathsf{H} \\
					& \quad + \sum_{g' \ge g} {\mathbf{H}_{g'}^\mathrm{B}}^\mathsf{H} \mathbf{H}_{g'}^\mathrm{B} \mathbf{\Theta}_{g'}^{(r)} \mathbf{H}_{g'}^\mathrm{F} {\mathbf{H}_{g'}^\mathrm{F}}^\mathsf{H}.
				\end{split}
			\end{equation}
		\end{proposition}

@@

		\begin{figure}[!t]
			\centering
			\resizebox{0.65\columnwidth}{!}{
				\input{assets/simulation/pc_power_sx.tex}
			}
			\caption{Average channel power versus \gls{ris} elements $N_\mathrm{S}$ and group size $L$ for $(N_\mathrm{T}, N_\mathrm{R}) = (8, 4)$, $(\Lambda^\mathrm{D}, \Lambda^\mathrm{F}, \Lambda^\mathrm{B}) = (65, 54, 46) \unit{dB}$.}
			\label{sm:pc_power_sx}
		\end{figure}

		\begin{figure}[!t]
			\centering
			\subfloat[Without Direct Link\label{sm:pc_power_bond_nd}]{
				\resizebox{0.48\columnwidth}{!}{
					\input{assets/simulation/pc_power_bond_nd.tex}
				}
			}
			\subfloat[With Direct Link\label{sm:pc_power_bond_hd}]{
				\resizebox{0.48\columnwidth}{!}{
					\input{assets/simulation/pc_power_bond_hd.tex}
				}
			}
			\caption{Average channel power versus \gls{ris} group size $L$ for $(N_\mathrm{T}, N_\mathrm{S}, N_\mathrm{R}) = (8, 256, 4)$, $(\Lambda^\mathrm{D}, \Lambda^\mathrm{F}, \Lambda^\mathrm{B}) = (65, 54, 46) \unit{dB}$.}
			\label{sm:pc_power_bond}
		\end{figure}
		% Fig.~\ref{sm:pc_power_sx} shows that, apart from numerous reflecting elements, a sufficiently large group size can help
		% Fig.~\ref{sm:pc_power_sx} shows that increasing the group size can improve the channel power especially for a large \gls{ris}.
		Fig.~\ref{sm:pc_power_sx} shows that, apart from adding reflecting elements $N_\mathrm{S}$, increasing the group size $L$ also improves the channel power.
		% The behavior is more obvious for a large \gls{ris}.
		This behavior is more pronounced for a large \gls{ris}.
		For example, the gain of pairwise connection is 2.8\% for $N_\mathrm{S} = 16$ and 28\% for $N_\mathrm{S} = 256$.
		% \gls{bd} \gls{ris} provides a higher channel power than conventional \gls{ris}.
		% As $L$ increases from 1 to 2,
		It implies that the channel shaping capability of \gls{bd} \gls{ris} scales with group size $L$.

@@

		% neither effectively utilizes the cascaded channel power, nor effectively align the direct-indirect eigenspace.
		% We observe that conventional \gls{ris} neither effectively utilizes the cascaded channel power, nor effectively align the direct-indirect eigenspace.
		% For conventional \gls{ris},
		% align the direct-indirect eigenspace.
		% When the direct link is present, a small $L$ can neither align the direct-indirect eigenspace nor the forward-backward eigenspace.
		% For a relatively large $N_\mathrm{S}$, conventional \gls{ris} cannot
		% neither the direct-indirect nor the forward-backward eigenspace can be effectively aligned with a small $L$.

@@

		% When the direct link is absent, a large $L$ can align the forward-backward eigenspace but cannot preserve the forward-backward power balance.
	\end{subsection}
\end{section}

>>>>

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}

@@

\begin{section}{Point-to-Point MIMO}
	\begin{subsection}{Channel Power Maximization}
		Consider a \gls{bd} \gls{ris} with $N^\mathrm{S}$ elements, which is divided into $G$ groups of equal $L$ elements.
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\left\lVert \mathbf{H}^\mathrm{D} + \sum_g\nolimits \mathbf{H}_g^\mathrm{B} \mathbf{\Theta}_g \mathbf{H}_g^\mathrm{F} \right\rVert _\mathrm{F}^2}{\label{op:pc_power}}{}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g \in \mathcal{G} \triangleq \{1,\ldots,G\}.}{}{}
		\end{maxi!}
		For \emph{symmetric} BD-RIS, the problem has been solved in
		\begin{itemize}
			\item Matteo's paper \cite{Nerini2023}: SISO and equivalent\footnote{Single-stream MIMO with given precoder and combiner.};
			\item Ignacio's paper \cite{Santamaria2023}: SISO and directless MISO/SIMO.
		\end{itemize}

		\begin{remark}
			% The difficulty of \eqref{op:pc_power} is that the \gls{ris} needs to balance the (additive) direct-indirect eigenspace alignment and the (multiplicative) forward-backward eigenspace alignment.
			The difficulty of \eqref{op:pc_power} is that the \gls{ris} needs to balance the additive (direct-indirect) and multiplicative (forward-backward) eigenspace alignment.
			% align the direct and indirect eigenspaces while preserving the
			Interestingly, it has the same form as the \emph{weighted orthogonal Procrustes problem} \cite{Gower2004}:
			\begin{mini!}
				{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{C} - \mathbf{A \Theta B} \rVert _\mathrm{F}^2}{\label{op:weighted_orthogonal_procrustes}}{}
				\addConstraint{\mathbf{\Theta}^\mathsf{H} \mathbf{\Theta}=\mathbf{I}.}{}{}
			\end{mini!}
			There exists no trivial solution to \eqref{op:weighted_orthogonal_procrustes}.
			One lossy transformation, by moving $\mathbf{\Theta}$ to one side \cite{Bell2003}, formulates a standard orthogonal Procrustes problem:
			\begin{mini!}
				{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{A}^\dagger \mathbf{C} - \mathbf{\Theta B} \rVert _\mathrm{F}^2}{\label{op:standard_orthogonal_procrustes}}{}
				\addConstraint{\mathbf{\Theta}^\mathsf{H} \mathbf{\Theta}=\mathbf{I}.}{}{}
			\end{mini!}
			\eqref{op:standard_orthogonal_procrustes} has a global optimal solution $\mathbf{\Theta}^\star = \mathbf{U} \mathbf{V}^\mathsf{H}$, where $\mathbf{U}$ and $\mathbf{V}$ are left and right singular matrix of $\mathbf{\mathbf{A}^\dagger \mathbf{C} \mathbf{B}^\mathsf{H}}$ \cite{Golub2013}.
			This low-complexity solution will be compared with the one proposed later.
		\end{remark}

		Inspired by \cite{Nie2017}, we propose an iterative algorithm to solve \eqref{op:pc_power}.
		The idea is to successively approximate the quadratic objective with a sequence of affine functions and solve the resulting subproblems in closed form.

		\begin{proposition}
			Start from any $\mathbf{\Theta}^{(0)}$, the sequence
			\begin{equation}
				\mathbf{\Theta}_g^{(r+1)} = \mathbf{U}_g^{(r)} \mathbf{V}_g^{(r)}, \quad \forall g \in \mathcal{G}
			\end{equation}
			converges to a stationary point of \eqref{op:pc_power}, where $\mathbf{U}_g^{(r)}$ and $\mathbf{V}_g^{(r)}$ are left and right singular matrix of
			\begin{equation}
				\begin{split}
					\mathbf{M}_g^{(r)}
					& = {\mathbf{H}_g^\mathrm{B}}^\mathsf{H} \mathbf{H}^\mathrm{D} {\mathbf{H}_g^\mathrm{F}}^\mathsf{H} + \sum_{g' < g} {\mathbf{H}_{g'}^\mathrm{B}}^\mathsf{H} \mathbf{H}_{g'}^\mathrm{B} \mathbf{\Theta}_{g'}^{(r+1)} \mathbf{H}_{g'}^\mathrm{F} {\mathbf{H}_{g'}^\mathrm{F}}^\mathsf{H} \\
					& \quad + \sum_{g' \ge g} {\mathbf{H}_{g'}^\mathrm{B}}^\mathsf{H} \mathbf{H}_{g'}^\mathrm{B} \mathbf{\Theta}_{g'}^{(r)} \mathbf{H}_{g'}^\mathrm{F} {\mathbf{H}_{g'}^\mathrm{F}}^\mathsf{H}.
				\end{split}
			\end{equation}
		\end{proposition}

@@

		\begin{figure}[!t]
			\centering
			\resizebox{0.65\columnwidth}{!}{
				\input{assets/simulation/pc_power_sx.tex}
			}
			\caption{Average channel power versus \gls{ris} elements $N^\mathrm{S}$ and group size $L$ for $(N^\mathrm{T}, N^\mathrm{R}) = (8, 4)$, $(\Lambda^\mathrm{D}, \Lambda^\mathrm{F}, \Lambda^\mathrm{B}) = (65, 54, 46) \unit{dB}$.}
			\label{sm:pc_power_sx}
		\end{figure}

		\begin{figure}[!t]
			\centering
			\subfloat[Without Direct Link\label{sm:pc_power_bond_nd}]{
				\resizebox{0.48\columnwidth}{!}{
					\input{assets/simulation/pc_power_bond_nd.tex}
				}
			}
			\subfloat[With Direct Link\label{sm:pc_power_bond_hd}]{
				\resizebox{0.48\columnwidth}{!}{
					\input{assets/simulation/pc_power_bond_hd.tex}
				}
			}
			\caption{Average channel power versus \gls{ris} group size $L$ for $(N^\mathrm{T}, N^\mathrm{S}, N^\mathrm{R}) = (8, 256, 4)$, $(\Lambda^\mathrm{D}, \Lambda^\mathrm{F}, \Lambda^\mathrm{B}) = (65, 54, 46) \unit{dB}$.}
			\label{sm:pc_power_bond}
		\end{figure}
		% Fig.~\ref{sm:pc_power_sx} shows that, apart from numerous reflecting elements, a sufficiently large group size can help
		% Fig.~\ref{sm:pc_power_sx} shows that increasing the group size can improve the channel power especially for a large \gls{ris}.
		Fig.~\ref{sm:pc_power_sx} shows that, apart from adding reflecting elements $N^\mathrm{S}$, increasing the group size $L$ also improves the channel power.
		% The behavior is more obvious for a large \gls{ris}.
		This behavior is more pronounced for a large \gls{ris}.
		For example, the gain of pairwise connection is 2.8\% for $N^\mathrm{S} = 16$ and 28\% for $N^\mathrm{S} = 256$.
		% \gls{bd} \gls{ris} provides a higher channel power than conventional \gls{ris}.
		% As $L$ increases from 1 to 2,
		It implies that the channel shaping capability of \gls{bd} \gls{ris} scales with group size $L$.

@@

		% neither effectively utilizes the cascaded channel power, nor effectively align the direct-indirect eigenspace.
		% We observe that conventional \gls{ris} neither effectively utilizes the cascaded channel power, nor effectively align the direct-indirect eigenspace.
		% For conventional \gls{ris},
		% align the direct-indirect eigenspace.
		% When the direct link is present, a small $L$ can neither align the direct-indirect eigenspace nor the forward-backward eigenspace.
		% For a relatively large $N^\mathrm{S}$, conventional \gls{ris} cannot
		% neither the direct-indirect nor the forward-backward eigenspace can be effectively aligned with a small $L$.

@@

		% When the direct link is absent, a large $L$ can align the forward-backward eigenspace but cannot preserve the forward-backward power balance.
	\end{subsection}

	\begin{subsection}{Rate Maximization}
		The problem is formulated w.r.t. precoder (instead of transmit covariance matrix) for reference:
		\begin{maxi!}
			{\scriptstyle{\mathbf{W},\mathbf{\Theta}}}{\log \det \Bigl(\mathbf{I} + \frac{\mathbf{W}^\mathsf{H}\mathbf{H}^\mathsf{H}\mathbf{H}\mathbf{W}}{\sigma_n^2}\Bigr)}{\label{op:pc_rate}}{}
			\addConstraint{\lVert \mathbf{W} \rVert _\mathrm{F}^2 \le P}{}{}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g.}{}{\label{cs:pc_rate_unitary}}
		\end{maxi!}

		\eqref{op:pc_rate} is jointly non-convex and solved by \gls{ao}.
		For a given $\mathbf{\Theta}$, the optimal precoder is given by
		\begin{equation}
			\mathbf{W}^\star = \mathbf{V} {\mathbf{S}^\star}^{1/2},
		\end{equation}
		where $\mathbf{V}$ is right singular matrix of $\mathbf{H}$ and $\mathbf{S}^\star$ is a diagonal matrix of the water-filling power allocation.
		For a given $\mathbf{W}$, we update $\mathbf{\Theta}$ by \gls{rcg} method along the geodesics \cite{Abrudan2009}.
		\begin{remark}
			A geodesic refers to the shortest path between two points in a Riemannian manifold.
			% For unitary constraint \eqref{cs:pc_rate_unitary}, the
			Unitary constraint \eqref{cs:pc_rate_unitary} translates to a Stiefel manifold where the geodesics have simple expressions described by the exponential map \cite{Abrudan2008}.
			% Geodesics are the locally shortest paths in a Riemannian manifold.
			% A Geodesics are the locally shortest paths in a Riemannian manifold.
			% on Riemannian manifold are the shortest path between two points.
		\end{remark}
		For optimization problems with block unitary constraint, the adapted \gls{rcg} method at iteration $r$ for block $g$ is summarized below.
		\begin{enumerate}
			\item Compute the Euclidean gradient
			\begin{equation}
				\nabla_{\mathbf{\Theta}_g}^\mathrm{E} f^{(r)} = \frac{\partial f^{(r)}}{\partial \mathbf{\Theta}_g^*}
			\end{equation}
			\item Translate to the Riemannian gradient
			\begin{equation}
				\nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r)} = \nabla_{\mathbf{\Theta}_g}^\mathrm{E} f^{(r)} {\mathbf{\Theta}_g^{(r)}}^\mathsf{H} - \mathbf{\Theta}_g^{(r)} {\nabla_{\mathbf{\Theta}_g}^\mathrm{E} f^{(r)}}^\mathsf{H}
			\end{equation}
			\item Determine the weight factor
			\begin{equation}
				% \gamma^{(r)} = \frac{\mathrm{tr}\Bigl(\bigl(\mathbf{G}_\mathrm{R}^{(r)} - \mathbf{G}_\mathrm{R}^{(r-1)}\bigr) \mathbf{G}_\mathrm{R}^{(r)\mathsf{H}}\Bigr)}{\mathrm{tr}\Bigl(\mathbf{G}_\mathrm{R}^{(r-1)} \mathbf{G}_\mathrm{R}^{(r-1)\mathsf{H}}\Bigr)}
				\gamma^{(r)} = \frac{\mathrm{tr}\Bigl(\bigl(\nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r)} - \nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r-1)}\bigr) \nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r)\mathsf{H}}\Bigr)}{\mathrm{tr}\Bigl(\nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r-1)} \nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r-1)\mathsf{H}}\Bigr)}
			\end{equation}
			\item Compute the conjugate direction
			\begin{equation}
				% \mathbf{D}_\mathrm{R}^{(r)} = \mathbf{G}_\mathrm{R}^{(r)} + \gamma^{(r)} \mathbf{D}_\mathrm{R}^{(r-1)}
				\mathbf{D}^{(r)} = \nabla_{\mathbf{\Theta}_g}^\mathrm{R} f^{(r)} + \gamma^{(r)} \mathbf{D}^{(r-1)}
			\end{equation}
			\item Determine the Armijo step size
			\begin{equation}
				\mu^{(r)} = \arg\max_\mu f^{(r)}\Bigl(\exp\bigl(-\mu \mathbf{D}^{(r)}\bigr) \mathbf{\Theta}_g^{(r)}\Bigr)
			\end{equation}
			\item Perform multiplicative (rotational) update
			\begin{equation}
				\mathbf{\Theta}^{(r+1)} = {\color{gray}\exp}\Bigl(-\mu^{(r)} \mathbf{D}^{(r)}\Bigr) \mathbf{\Theta}_g^{(r)}
			\end{equation}
		\end{enumerate}
		The advantage over general manifold optimization \cite{Absil2009,Pan2022d} is trifold: 1) no retraction is required; 2) no projection is required; 3) no matrix inversion is required.
	\end{subsection}
\end{section}
